---
title: "Introduction to docstring"
author: "Dason Kurkiewicz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to docstring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


The docstring package is a package for R that provides the ability to 
display something analagous to
Python's docstrings within R.  By allowing the user to document
their functions as comments at the beginning of their function
without requiring putting the function into a package we allow
more users to easily provide documentation for their functions.
The documentation can be viewed using an accessor function but
displays just like any other R help files.

The user will need to be familiar with roxygen style comments (via the [roxygen2 package](https://cran.r-project.org/package=roxygen2))
to fully utilize the package.

In this vignette we will walk through a few examples to give you an introduction
to what the package does and some simple examples of how to format your
documentation to be used with docstring.


```{r, include=FALSE}
options(help_type="text")

# A pager that outputs to the console
console_pager <- function(x, ...){
    input <- readLines(x)
    # Had some issues with _ getting displayed
    # in the output console output which
    # messed up rendering in the created html vignette
    # So remove that before outputting.
    input <- gsub("_", "", input)
    cat(paste(input,collapse="\n"), "\n")}
options(pager=console_pager)
options(docstring_rstudio_help_pane = FALSE)
```

At it's most basic documentation can just be a sentence or two describing
what a function does.

```{r}
square <- function(x){
    #' Computes the square of the input
    return(x^2)
}
```

The comment at the beginning serves as our documentation of what the function does.
If we wanted to view the documentation we could print the entire function
in which case we would be able to see the documentation in the source.

```{r, echo=TRUE}
square
```

However for longer functions that can be incovenient. The built in help system in
R is the ideal way to view help for functions.  By default we can't use the 
help system to view the documentation we wrote in the `square` function.

```{r}
?square
```

However, using the docstring package will allow us to do just that for functions
that have a corresponding docstring.  For functions that don't have a docstring
it will fall back to the 'typical' functionality of `?`. Note that this vignette
has been configured to display the help documentation as output in the console.

```{r}
library(docstring)
```
```{r, comment=NA}
?square # Below is the generated help file:
```

## More complex docstring

When the docstring is just a single chunk of text you'll notice that the
generated help file puts the text in the "Description" section and has
"Title Not Detected". This is a convenience feature but the documentation comments
can (and should in most cases) follow the standards set by the roxygen2 package.
So this means that the first chunk of comments before a blank line is taken to
be the title, the second chunk is taken to be the Description section, and the
rest goes in the "Details" section.  You can have finer control over these things
by using keywords but let's look at an example in action

```{r, comment=NA}
test <- function(){
    #' This is my title line
    #'
    #' All of this text goes
    #' in the Description section
    #'
    #' This part goes in the Details!
    return()
}

?test
```

You'll notice that the "usage" portion gets generated automatically.

### Using keywords

The standard set in roxygen2 allows the use of keywords to specify different
sections and the corresponding documentation.  For example if you want to
provide documentation explaining what one of the input parameters means you
could include a line that starts with `@param parameter_name` which will 
create an Arguments section and place the documentation about the input parameters
there.

```{r, comment=NA}
square <- function(x){
    #' Squares a number
    #'
    #' Provides the square of the input
    #' @param x The value to be squared
    return(x^2)
}

?square
```

You can see that the "Arguments" section gets automatically generated using
the documentation you wrote on the line that started with `@param`.  When using
docstring it isn't expected that you are in the process of building a package.
If you were you would most likely put your documentation above the function as
that is the style that is typically required in package building via `roxygen2`.
Due to that I'll only describe a subset of the keywords that are available to
the user but if you are interested in more on how elaborate you can make your
documentation I would suggest reading the [Generating Rd files](https://cran.r-project.org/package=roxygen2/vignettes/rd.html) vignette
for the roxygen2 package.

Some of the other useful keywords that I can forsee users using with docstring are:

 - `@return description` to describe the output that a function creates.
 - `@title`, `@description`, `@details` are available if you want to be more
 explicit in the sections. This also allows you to have a description section
 longer than one paragraph if you explicitly use the description keyword.
 - `@note` Creates a note section
 - `@section SectionTitle:` You can create your own sections using the section
 keyword.  Following `@section` must be the section title in sentence case and
 must be finished with a colon. Subsections do not have their own keyword but 
 can be added using the Rd `\subsection{}` command.
 - `@usage` If you want to have more control over the usage section you can
 either overwrite the generated usage or add to it by using this keyword
 - `@references` If you want to add references to your documentation. This is
 important (and appreciated) especially when you are writing a function that
 implements an algorithm in a journal article.  Having a reference to the source
 is vital in those situations.
 - `@examples` which will create an examples section in the documentation. Using
 docstring there isn't a nice way to run the examples like you can using the
 `example()` function but they will still show up in the documentation which
 is sometimes the most important part. Users can still copy/paste if they desire.
 - `@export`, `@import`, `@importFrom` These are keywords that are highly useful
 for and only really make sense in the context of package creation.  docstring
 isn't meant to be used for full packages so you wouldn't expect me to comment
 on these keywords. However, docstring is a nice tool to use in the time before
 package creation so it is understandable that if somebody thought their
 code will turn into a package someday that they might want to get a jump start
 on *all* of the documentation (including the namespace directives).  There is
 nothing stopping a user from using these keywords.  They won't do anything
 when used with docstring though.
 
 So now to illustrate all of this I will present an example function that uses
 **all** of these keywords.  The function is pointless and the documentation is
 gibberish but it should allow you to see how to format the docstring using
 the keywords and what the corresponding help file will look like.

```{r, comment=NA}
mypaste <- function(x, y = "!"){
  #' Paste two items
  #' 
  #' @description This function pastes two items
  #' together.  
  #'
  #' By using the description tag you'll notice that I
  #' can have multiple paragraphs in the description section
  #' 
  #' @param x character. The first item to paste
  #' @param y character. The second item to paste Defaults to "!" but
  #' "?" would be pretty great too
  #' @usage mypaste(x, y)
  #' @return The inputs pasted together as a character string.
  #' @details The inputs can be anything that can be input into
  #' the paste function.
  #' @note And here is a note. Isn't it nice?
  #' @section I Must Warn You:
  #' The reference provided is a good read.
  #' \subsection{Other warning}{
  #'   It is completely irrelevant to this function though.
  #' }
  #' 
  #' @references Tufte, E. R. (2001). The visual display of 
  #' quantitative information. Cheshire, Conn: Graphics Press.
  #' @examples
  #' mypaste(1, 3)
  #' mypaste("hey", "you")
  #' mypaste("single param")
  #' @export
  #' @importFrom base paste

  return(paste(x, y))
}

?mypaste
```


